---
title: "allocation_water_filling"
author: "Konilo Zio"
date: "`r Sys.time()`"
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    papersize: a4
execute:
    echo: false
---

```{r}
#| label: setup
#| echo: false
#| output: false

# record t0 (see the appendix)
t0 <- proc.time()

# required packages
library(data.table)
library(ggplot2)
library(knitr)

# Runs when executing R interactively (because /workspaces/sandbox/ is the workdir),
# but not when rendering the qmd (because the qmd's dir is the workdir in that case)
if (getwd() == "/workspaces/sandbox") {
    setwd("sandbox/allocation_water_filling")
    library(httpgd)
    httpgd::hgd()
    # Plots will be displayed via a web page whose URL will be printed be the function above
}
```


# Introduction

When (i) investing regularly (e.g., monthly) (ii) following a strategy that involves more than one asset -- each with its own target allocation -- while (iii) ruling out asset selling, computing the amount to invest in each asset at a point in time can be tricky.


# Simplest case

Let's say we just received our pay and now have 1 000 € to invest.

Let's say that, at that point, the portfolio is like so:

| Asset | Target allocation (%) | Actual allocation (%) | Target value (€) | Actual value (€) |
|---|---|---|---|---|
| Index ETF |  90 % |  95 % |  9 000 € |  9 500 € |
| Managed futures ETF |  10 % |   5 % |  1 000 € |   500 € |
| Total portfolio | 100 % | 100 % | 10 000 € | 10 000 |
: Pre-investment situation {#tbl-1-pre}

Adding 1 000 € to the portfolio would lead to the following value targets

| Asset | Target allocation | Target value | Pre-investment value | Delta |
|---|---|---|---|---|
| Index ETF | 90 % | 9 900 € | 9 500€ | + 400 € |
| Managed futures ETF | 10 % | 1 100 € | 1 000 € | + 600 € |
| Total portfolio | 100 % | 11 000 € | 10 000 € | + 1000 € |
: Post-investment targets {#tbl-1-post}

In that case, 400 € of the 1000 € investment are allocated to the index ETF and 600 € to the managed futures ETF to reach the targets. The math is simple, no selling is required.


# When selling is allowed

Now, let's say that just before investing, the portfolio is like so:

| Asset | Target allocation | Actual allocation | Target value | Actual value |
|---|---|---|---|---|
| Index ETF | 90 % | 95 % | 27 000 € | 28 500 € |
| Managed futures ETF | 10 % | 5 % | 3 000 € |  1 500 € |
| Total portfolio | 100 % | 100 % | 30 000 € | 30 000 |
: Pre-investment situation {#tbl-2-pre}

Adding 1 000 € to the portfolio would lead to the following value targets

| Asset | Target allocation | Target value | Pre-investment value | Delta |
|---|---|---|---|---|
| Index ETF | 90 % | 27 900 € | 28 500 € | -   600 € |
| Managed futures ETF | 10 % |  3 100 € |  1 500 € | + 1 600 € |
| Total portfolio | 100 % | 31 000 € | 30 000€ | + 1 000 € |
: Post-investment targets {#tbl-2-post}

In that case you can see that, to reach the targets, we would need to
- sell 600 € worth of index ETF shares
- and buy 1 600 € of managed futures ETF shares (i.e., the 1 000 € investment + 600 € of cash from selling the index ETF shares).

That's still simple, mathemically. But, from a practical standpoint, selling is generally avoided if possible in order to minimize order fees which can substantially damage the portfolio's return if repeated (x% of the order's amount is paid by the investor).


# When selling is prohibited -- simplest case

We'll consider the same pre-investment situation as above. But, this time, the investor chooses to prohibit selling -- in order to avoid selling order fees -- because they think they contribute enough every month to keep their portfolio approximately balanced.

In that case, they could simply invest all their contribution in the underweight asset and be done. @tbl-3-post shows that approach would reduce the managed futures ETF's delta while leaving the index ETF's delta intact.

| Asset | Target allocation | Target value | Actual value | Delta |
|---|---|---|---|---|
| Index ETF | 90 % | 27 900 € | 28 500 € | -   600 € |
| Managed futures ETF | 10 % | 3 100 € |  2 500 € | +   600 € |
| Total portfolio | 100 % | 31 000 € | 31 000 € | +     0 € |
: Post-investment situation {#tbl-3-post}

TODO measure pre-invest imbalance.


# When selling is prohibited -- when it gets tricky

Now, what if we have 5 000 € to invest? Using the naive apporach from above would yield the situation in @tbl-4-post: the portfolio goes from managed futures ETF-underweight to index ETF-underweight and, most importantly for our strategy, it gets more unbalanced as the imbalance goes from __ to 3 000 €

| Asset | Target allocation | Target value | Actual value | Delta |
|---|---|---|---|---|
| Index ETF | 90 % | 31 500 € | 28 500 € | + 3 000 € |
| Managed futures ETF | 10 % |  3 500 € |  6 500 € | - 3 000 € |
| Total portfolio | 100 % | 35 000 € | 35 000 € | +     0 € |
: Post-investment situation {#tbl-4-post}

In that case, we need another, two step, mathematical approach to compute how to allocate the 1 000 €.


# When selling is prohibited and there are 2+ underweight assets -- when it gets actually laborious


# Draft

The breakeven between the two scenarios involves the following variables:

- $\text{days}$ = inter-trade duration in trading days
- $\text{daily return}$ = expected net daily return of the risk-off asset
- $\text{order fee}$ = order fee applied to buy and sell orders by the broker
- $\text{cash}$ = initial amount of liquidities (cash) to manage with either option

@eq-break models the problem and finds the formula to compute the number of \text{days} necessary to breakeven.
$$
\text{cash} = \text{cash} (1 - \text{order fee}) (1 + \text{daily return})^\text{days} (1 - \text{order fee})
$$ {#eq-break}



blabla
```{r}
print("yes")
```



# Appendix

This `qmd` took `r round((proc.time() - t0)[3] / 60)` minutes to render. It was rendered in the following environment:
```{r}
#| label: appendix

print(sessionInfo(), locale = FALSE) |>
    capture.output() |>
    strwrap(width = 70, simplify = FALSE) |>
    unlist() |>
    cat(sep = "\n")
```
